A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-18 paper.tex
Added a few clarifying notes

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
% TODO do we mention "marginal" anywhere? Might not be worth the
% effort, surely "local" is clear?
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

@@

% % The full ancestral recombination graph (ARG) is a structure that encodes all
% % coalescence and recombination events resulting from the stochastic process of
% % the coalescent with recombination.

% historial trend discussed in the XXX section).
% \citet{rasmussen2022espalier} have a less restrictive definition, in
% which a ``full'' ARG is one that contains complete information about
% recombination events, without necessarily being tied to a particular
% stochastic process.
% While [something positive], it is important to be clear on
% what the overall goal of such precise inferences are,
% what the limitations on when they can be made are,
% and their downstream utility in applications.

In an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
``unary'' nodes in the local trees
(where a node has exactly one child over a particular genomic region).
% "unary regions" is jargon - just dropping it in here like this doesn't
% clarify. We can't assume that people are used to switching between
% graphs and trees.
% of ``unary'' regions for nodes
% (where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.

>>>>

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
%

@@

In the classical view of an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
``unary'' nodes in the local trees
(where a node has exactly one child over a particular genomic region).
% "unary regions" is jargon - just dropping it in here like this doesn't
% clarify. We can't assume that people are used to switching between
% graphs and trees.
% of ``unary'' regions for nodes
% (where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citep{mathieson2020ancestry} show examples in which the ARGs 
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{minichiello2006mapping} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]
